---
title: "Amplicon Sorting code"
AI assistance provided by Claude, Sonnet v4.6 (Anthropic, 2026). https://claude.ai
date: "02/26-04/26"
output:
---

Loading libraries

```{r}

library(tidyverse)
library(tidytext)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE) #keeps these packages installed for all future chunks
options(scipen = 999)
```
Loading data in

```{r}
plate_reader <- function(plate){
  
  path <- paste("csv_files/", plate, sep="") #define path
  input_length <- list.files(path=path) #define how long the list is
  
  for(i in 1:length(input_length)) { #looping through all the files in the dir
    current.var <- paste(path, input_length[i], sep="") #define the file, as whatever step the loop is on within this specific path
    file_name <- read.csv(current.var) #read the .csv file into a var
    assign(input_length[i], file_name, envir = .GlobalEnv) #assign the read csv file to the global environment with the correct name
  
    }
}

plate_reader("Lakes_1/")
plate_reader("Lakes_2/")
plate_reader("Lakes_3/")
plate_reader("BW_1/")
plate_reader("BW_2/")
plate_reader("BW_3/")



```
Cleaning the data up:
Checking that all columns match with one another and combining into one huge data frame.

Col name checking:
```{r}

#Checking that all the columns match up with each other:

#This allows us to see if there are any non-matching headers between the data frames that may cause issues
#BW_1

a <- names(BW_1_18S.csv)
b <- names(BW_1_28S.csv)
c <- names(BW_1_Moorea.csv)
d <- names(BW_1_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)

#BW_2
a <- names(BW_2_18S.csv)
b <- names(BW_2_28S.csv)
c <- names(BW_2_Moorea.csv)
d <- names(BW_2_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)
#all good there

#BW_3
a <- names(BW_3_18S.csv)
b <- names(BW_3_28S.csv)
c <- names(BW_3_Moorea.csv)
d <- names(BW_3_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)
#all good there

#Lakes 1
a <- names(Lakes_1_18S.csv)
b <- names(Lakes_1_28S.csv)
c <- names(Lakes_1_Moorea.csv)
d <- names(Lakes_1_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)

#Lakes 2
a <- names(Lakes_2_18S.csv)
b <- names(Lakes_2_28S.csv)
c <- names(Lakes_2_Moorea.csv)
d <- names(Lakes_2_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)
#all good

#Lakes 3
a <- names(Lakes_3_18S.csv)
b <- names(Lakes_3_28S.csv)
c <- names(Lakes_3_Moorea.csv)
d <- names(Lakes_3_Sauron.csv)
setdiff(a, b)
setdiff(a, c)
setdiff(a, d)
setdiff(b, c)
setdiff(b, d)
setdiff(c, d)
#all good
```

Combining all of the separate plate files into a single HUGE data frame

```{r}
csv_names <- ls(pattern = ".csv") #gets all the names of objects ending in .csv
csv_list <- mget(csv_names) #compiles everything into a list


#this deals with an issue between the data frames. Some hit2_group columns are integers whilst some are characters. This forces them all to be chr
csv_data_list <- lapply(csv_list, function(df) {
  df %>% mutate(hit2_group = as.character(hit2_group))
})

#bind all the data frames together into a big one
bigboi <- bind_rows(csv_data_list)
bigboi <- bigboi[-c(1713, 2098), ]#remove two erroneous rows

#Correction of 2 mislabelled samples
bigboi <- bigboi %>%
  mutate(expected_taxon = case_when(
    plate == "Lakes_2" & barcode == "CO1" & sample == "blast" ~ NA,
    TRUE ~ expected_taxon
  ))
bigboi <- bigboi %>%
  mutate(expected_taxon = case_when(
    plate == "Lakes_1" & sample == "Accidental extraction blank. Pesky snails" ~ NA,
    TRUE ~ expected_taxon
  ))
#standardising names
bigboi <- bigboi %>% mutate(barcode = str_replace(barcode, "CO1", "COI"))
bigboi <- bigboi %>% mutate(plate = str_replace(plate, "BW_1", "Gardens_1"))
bigboi <- bigboi %>% mutate(plate = str_replace(plate, "BW_2", "Gardens_2"))
bigboi <- bigboi %>% mutate(plate = str_replace(plate, "BW_3", "Gardens_3"))
bigboi <- bigboi %>% mutate(hit1_expect = str_replace(hit1_expect, "y", "Y"))
bigboi <- bigboi %>% mutate(hit2_expect = str_replace(hit2_expect, "y", "Y"))

```

Sorting a sample list across all taxa and plotting stacked bars.

To do this, only a single plate of barcodes per analysed plate is required. rbind all together, then plot stacked bar.

```{r}

all_18S <- subset(bigboi, barcode == "18S") #select only the 18S plate - choice of plate is arbitrary
all_taxa_list <- all_18S[, c("plate", "expected_taxon")] #select only the plate and expected taxon

#Checking sample numbers
counting <- all_taxa_list %>%
  count(plate)

plot_data <- all_taxa_list %>%
  group_by(plate) %>% #group taxa by plate they belong to
  count(expected_taxon) %>% #count all of the expected taxa
  mutate(relative_abundance = n / sum(n) * 100) #work out the % share of each within the plate

ggplot(plot_data, aes(x = plate, y = relative_abundance, fill = expected_taxon)) +
  geom_bar(stat = "identity") +
  xlab("Dataset") +
  ylab("Percentage of samples") +
  scale_fill_manual(name = "Taxa", breaks = c("Acari", "Annelida", "Ciliophora", "Cladocera", "Copepoda", "Gastrotricha", "Insecta", "Mollusca", "Nematoda", "Ostracoda", "Other Protista", "Platyhelminthes", "Rotifera", "Tardigrada", "NA"), 
                       values=c("#E6A817", "#7B2D8B", "#FFEC8B", "#FF8C00", "#2DB34A", "#00C9A7", "#D4E84A", "#5B4FCF", "#E8334A", "#8B1A1A", "#C4956A", "#3A7A5C", "#FF6EC7", "#4A90D9", "#D6D6D6" )) +
  theme_minimal()



```

Building a unified data frame with success metrics:

```{r}

#Building a 'success' data frame to work with further down the line
success_metric <- bigboi[, c("file_status", "plate", "barcode", "expected_taxon", "hit1_expect", "hit2_expect", "final_expect", "hit1_primer_set", "hit2_primer_set")]

#Trimming out non-meiofaunal taxa
success_metric <- success_metric[!(success_metric$expected_taxon == "Mollusca"), ]
success_metric <- success_metric[!(success_metric$expected_taxon == "Other Protista"), ]
success_metric <- success_metric[!(success_metric$expected_taxon == "Insecta"), ]
success_metric <- success_metric %>% filter(!is.na(expected_taxon))



```

```{r}
#############STACKED BARPLOT INCLUDING hit1/2############

plot_data_hit_inc <- success_metric %>%
  mutate(result_type = case_when( #case_when to add a single "result_type" column
    hit1_expect == "Y" ~ "MRC BLAST match", #Max_readcount was the blast match
    hit2_expect == "Y" ~ "AC BLAST match", #Other contig was the origin of the blast match
    file_status == "Y" & final_expect == "N" ~ "Off target/No Match", #none of the examined contigs matched the expected taxa
    file_status == "N" ~ "No contig present" #no file
    
  )) %>%
  filter(!is.na(result_type)) %>% #get rid of any NAs
  mutate(result_type = factor(result_type, 
                              levels = c("No contig present", "Off target/No Match",  "AC BLAST match", "MRC BLAST match"))) %>% #Stack the data properly for the graph
  group_by(result_type, barcode) %>% #pick out only the result_type, barcode and plate, count the result type for each unique barcode/plate combination
  count(result_type, barcode) %>%
  mutate(n) #work out the % share of each within the plate

# get percentage statistics for each barcode
plot_data_hit_barcode_stats <- plot_data_hit_inc %>%
  group_by(barcode) %>%
  mutate(percentB = n / sum(n) * 100)

# get percentage statistics for each result type
plot_data_hit_result_stats <- plot_data_hit_inc %>%
  group_by(result_type) %>%
  summarise(total_n = sum(n)) %>%
  mutate(percentR = (total_n / sum(total_n)) * 100)

#Gene plot
ggplot(plot_data_hit_inc, aes(x = barcode, y = n, fill = result_type)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Gene") +
  ylab("Proportion of potential contigs") +
  theme_minimal() +
  scale_fill_manual(name = "Sample Outcome", breaks = c("No contig present", "Off target/No Match",  "AC BLAST match", "MRC BLAST match"), values = c("grey40", "grey15", "darkolivegreen3", "#008F7C"))

plot_data_hit_inc_taxa <- success_metric %>%
  mutate(result_type = case_when( #case_when to add a single "result_type" column
    hit1_expect == "Y" ~ "MRC BLAST match", #Max_readcount was the blast match
    hit2_expect == "Y" ~ "AC BLAST match", #Other contig was the origin of the blast match
    file_status == "Y" & final_expect == "N" ~ "Off target/No Match", #none of the examined contigs matched the expected taxa
    file_status == "N" ~ "No contig present" #no file
    
  )) %>%
  filter(!is.na(result_type)) %>% #get rid of any NAs
  mutate(result_type = factor(result_type, 
                              levels = c("No contig present", "Off target/No Match",  "AC BLAST match", "MRC BLAST match"))) %>% #Stack the data properly for the graph
  group_by(result_type, expected_taxon) %>% 
  count(result_type, expected_taxon) %>%
  mutate(n)

#taxa plot
ggplot(plot_data_hit_inc_taxa, aes(x = expected_taxon, y = n, fill = result_type)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Taxon") +
  ylab("Proportion of potential contigs") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
  scale_fill_manual(name = "Sample Outcome", breaks = c("No contig present", "Off target/No Match",  "AC BLAST match", "MRC BLAST match"), values = c("grey40", "grey15", "darkolivegreen3", "#008F7C"))


```


```{r}
############BREAKDOWN OF TAXA###################
plot_data_taxa <- success_metric %>%
  mutate(primer = case_when(
    hit1_expect == "Y" ~ hit1_primer_set,
    hit2_expect == "Y" ~ hit2_primer_set,
    file_status == "Y" & final_expect == "N" ~ "Off target/No Match", #none of the examined contigs matched the expected taxa
    file_status == "N" ~ "No contig present", #no file
    TRUE ~ NA
  )) %>%
  filter(!is.na(primer)) %>%
  filter(!is.na(expected_taxon)) %>%
  mutate(primer = factor(primer, 
                              levels = c("No contig present", "Off target/No Match", "18S+", "28S", "Moorea", "Sauron"))) %>%
  group_by(barcode, expected_taxon) %>% 
  count(primer) %>%
  mutate(relative_abundance = n)

plot_data_taxa <- plot_data_taxa[!is.na(plot_data_taxa$primer), ] #removes any NA values in "primer"

###STATS###
#All taxa, specific stats as to where each Y has come from
plot_data_taxa_barcode_stats <- plot_data_taxa %>%
  group_by(barcode, expected_taxon) %>%
  mutate(percentB = n / sum(n) * 100)

#taxa overall Y/N stats
taxa_success <- success_metric %>%
  mutate(primer = case_when(
    hit1_expect == "Y" ~ "Y",
    hit2_expect == "Y" ~ "Y",
    file_status == "Y" & final_expect == "N" ~ "Off target/No Match", #none of the examined contigs matched the expected taxa
    file_status == "N" ~ "No contig present", #no file
    TRUE ~ NA
  )) %>%
  filter(!is.na(primer)) %>%
  filter(!is.na(expected_taxon)) %>%
  group_by(expected_taxon, primer) %>%
  count(primer) %>%
  ungroup() %>%
  group_by(expected_taxon) %>%
  mutate(percentB = n / sum(n) * 100)


test <- subset(plot_data_taxa, expected_taxon == "Acari") %>%
  group_by(barcode, primer) %>%
  count(primer)

ggplot(plot_data_taxa, aes(x = expected_taxon, y = relative_abundance, fill = primer)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Taxon") +
  ylab("Proportion of potential contigs") +
  theme_minimal() +
  facet_wrap(~barcode) +
  scale_fill_manual(name = "Contig origin", breaks = c("No contig present", "Off target/No Match", "18S+", "28S", "Moorea", "Sauron"), values = c("grey40", "grey15", "maroon4", "lightskyblue", "olivedrab1", "darkolivegreen")) +
  coord_flip()


```



```{r}
#mean readcount statistics for each amplicon

bigboi_stats <- bigboi[!(bigboi$expected_taxon == "Mollusca"), ]
bigboi_stats <- bigboi_stats[!(bigboi_stats$expected_taxon == "Other Protista"), ]
bigboi_stats <- bigboi_stats[!(bigboi_stats$expected_taxon == "Insecta"), ]
bigboi_stats <- bigboi_stats %>% filter(!is.na(expected_taxon))

readcount_stats <- bigboi_stats %>%
  mutate(successful_contig_readcount = case_when( 
    hit1_expect == "N" ~ max_readcount, 
    hit2_expect == "N" ~ hit2_readcount, 
    file_status == "Y" & final_expect == "Y" ~ NA, 
    file_status == "N" ~ NA 
    
  )) %>%
  filter(!is.na(successful_contig_readcount)) #get rid of any NAs
 
mean(readcount_stats$successful_contig_readcount)

#18S+

readcount_stats_18S <- bigboi_stats %>%
  mutate(successful_contig_readcount = case_when( 
    hit1_expect == "Y" & hit1_primer_set == "18S+" ~ max_readcount, 
    hit2_expect == "Y" & hit1_primer_set == "18S+" ~ hit2_readcount, 
    file_status == "Y" & final_expect == "Y" ~ NA, 
    file_status == "N" ~ NA,
    TRUE ~ NA
  )) %>%
  filter(!is.na(successful_contig_readcount))
mean(readcount_stats_18S$successful_contig_readcount)

#28S
readcount_stats_28S <- bigboi_stats %>%
  mutate(successful_contig_readcount = case_when( 
    hit1_expect == "Y" & hit1_primer_set == "28S" ~ max_readcount, 
    hit2_expect == "Y" & hit1_primer_set == "28S" ~ hit2_readcount, 
    file_status == "Y" & final_expect == "Y" ~ NA, 
    file_status == "N" ~ NA,
    TRUE ~ NA
  )) %>%
  filter(!is.na(successful_contig_readcount))
mean(readcount_stats_28S$successful_contig_readcount)

#Moorea
readcount_stats_M <- bigboi_stats %>%
  mutate(successful_contig_readcount = case_when( 
    hit1_expect == "Y" & hit1_primer_set == "Moorea" ~ max_readcount, 
    hit2_expect == "Y" & hit1_primer_set == "Moorea" ~ hit2_readcount, 
    file_status == "Y" & final_expect == "Y" ~ NA, 
    file_status == "N" ~ NA,
    TRUE ~ NA
  )) %>%
  filter(!is.na(successful_contig_readcount))
mean(readcount_stats_M$successful_contig_readcount)

#sauron
readcount_stats_s <- bigboi_stats %>%
  mutate(successful_contig_readcount = case_when( 
    hit1_expect == "Y" & hit1_primer_set == "Sauron" ~ max_readcount, 
    hit2_expect == "Y" & hit1_primer_set == "Sauron" ~ hit2_readcount, 
    file_status == "Y" & final_expect == "Y" ~ NA, 
    file_status == "N" ~ NA,
    TRUE ~ NA
  )) %>%
  filter(!is.na(successful_contig_readcount))
mean(readcount_stats_s$successful_contig_readcount)
```

```{r versioning}
sessionInfo()
```
