---
title: "Adapter usage: thresholding"
author: "Srishti Arya"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    df_print: paged
    code_fold: show
---

# Intro

-   This is a review of the demultiplexing process
-   Prior investigation suggested that perhaps the issue lies in a lax error rate threshold in demuxing, and possibly a problem with rescued reads from pychopper containing primers in the middle of reads
-   Here, I tested 4 options
    -   default (error rate 0.2 and pychopped rescued reads included)
    -   lower error rate (to error rate 0.1)
    -   no rescued reads
    -   no rescued reads AND a lower demux error rate

# Setup

```{r setup, message=FALSE}
knitr::opts_chunk$set() # keep packages in knits
library(tidyverse)
library(readr)
library(ggplot2)
library(vioplot)
library(ggrepel)
library(patchwork)

options(scipen = 999) # doesn't use standard form 
```

# Adapter searches: are the adapters present in the right places?

## data wrangling: full adapter search

```{r load full adapter search data}
full<- 
  read_csv("../../../sense_check_adapter_primer/adapter_testing_thresholds/collated_full_presence.csv", col_names = FALSE) %>% # read in the csv
  as.data.frame() %>%  # make it into a data frame
  dplyr::rename("adapter" = X1, "file" = X2, "fwd" = X3, "rc" = X4, "hits" = X5) %>% 
  dplyr::mutate( # adding a column
    threshold = str_extract(file, "(?<=SP5_[0-9]{3}_)(.*?)(?=_day1)"), # which threshold?
    adapter_in_file = mapply(grepl, adapter, file), # is the adapter in this file?
    match = case_when( # a clean alternative to a nested ifelse
      hits > 0 & !adapter_in_file ~ "misassignment", # are there non-null hits of the adapter in the WRONG file?
      hits > 0 & adapter_in_file ~ "ok", # are there non-null hits of the adapter in the RIGHT file?
      TRUE ~ "ok")) # and if no hits then also ok since this can be the adapters matching the wrong file so it's ok if there are 0

full_summary<-
  full %>% 
  mutate(hits = ifelse(is.na(hits) | is.nan(hits), 0, hits)) %>%  # Coerce NA/NaN to 0
  group_by(threshold, adapter, match) %>%
  summarise(total_hits = sum(hits, na.rm = TRUE), .groups = "drop") %>% # counting hits per adapter)
  complete(threshold, adapter, match, fill = list(total_hits = 0)) %>% 
  group_by(threshold, adapter) %>%
  mutate(percent = 100 * total_hits / sum(total_hits, na.rm = TRUE)) %>% # make a % count
  ungroup() %>%
  dplyr::filter(threshold == "lowererate_norescue") # some testing was done locally when deciding on cleaning/demuxing params

```

## plotting the full adapter sequences' correct hits and misassigned hits

```{r plot full adapter search data, fig.height=5, fig.width=10}
full %>% 
  ggplot(aes(x = adapter, y = hits, fill = match))+
  geom_bar(stat = "identity")+
  geom_text(data = full_summary %>% dplyr::filter(match == "misassignment" | percent == 0),
            aes(x = adapter, y = 100, label = paste0(signif(percent, 1) %>% replace(is.na(.), 0), "%")),
            vjust = -1, hjust = 0.1, size = 3, angle = 45) +
  theme_minimal()+
  theme(axis.text.x = element_text(hjust=1, vjust=1, angle=45),
        legend.position = "bottom")
```

```{r plot full adapter search summary, fig.height=7, fig.width=5}
# with percentage
full_summary %>% 
  ggplot(aes(x = adapter, y = percent, fill = match)) +
  geom_bar(stat = "identity") +
  coord_flip()+
  geom_text(data = full_summary %>% dplyr::filter(match == "misassignment" | percent==0),
            aes(x = adapter, y = 100, label = paste0(signif(percent, 1)%>% replace(is.na(.), 0), "%")),
            vjust = 0, hjust = -0.1, size = 3) +
  expand_limits(y=c(0, 115))+
  labs(x = "Adapter",
       y = "Percentage of total hits",
       fill = "Assignment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, fig.height=5, fig.width=5}
full_summary %>%
  dplyr::filter(match == "misassignment") %>% 
  ggplot(aes(x = match, y = adapter, fill = percent)) +
  geom_tile() +
  geom_text(aes(label = paste0(signif(percent, 1), "%")), size = 3) +
  scale_fill_gradient(low = "white", high = "#bd465e", na.value = "#d3d3d3") +
  labs(x = "Match type", y = "Adapter", fill = "% Misassigned")
```

